# Association tests

Since TOPMed has many studies with related participants, we focus on linear mixed models. Logistic mixed models are also possible using GENESIS, see the [GMMAT paper](https://www.ncbi.nlm.nih.gov/pubmed/27018471).

## Null model

The first step in an association test is to fit the null model. We use the `AnnotatedDataFrame` with phenotypes, and a GRM. If the sample set involves multiple distinct groups with different variances for the phenotype, we recommend allowing the model to use heterogeneous variance among groups with the parameter `group.var`.

```{r null_model}
data.path <- "https://github.com/smgogarten/analysis_pipeline/raw/devel/testdata"
sampfile <- "1KG_phase3_subset_annot.RData"
if (!file.exists(sampfile)) download.file(file.path(data.path, sampfile), sampfile)
annot <- TopmedPipeline::getobj(sampfile)

grmfile <- "grm.RData"
if (!file.exists(grmfile)) download.file(file.path(data.path, grmfile), grmfile)
grm <- TopmedPipeline::getobj(grmfile)
# the row and column names of the covariance matrix must be set to sample.id
rownames(grm$grm) <- colnames(grm$grm) <- grm$sample.id

library(GENESIS)
nullmod <- fitNullMM(annot, outcome="outcome", covars=c("sex", "Population"), 
                     covMatList=grm$grm, group.var="Population", verbose=FALSE)
```

We also recommend taking an inverse normal transform of the residuals and refitting the model. This is done separately for each group, and the transformed residuals are rescaled. See the full procedure in the  
[pipeline documenation](https://github.com/smgogarten/analysis_pipeline#association-testing).

## Single-variant tests

Single-variant tests are the same as in GWAS. We use the `assocTestMM` function in GENESIS. We have to create a `SeqVarData` object including both the GDS file and the sample annotation containing phenotypes.

```{r assoc_single}
library(SeqVarTools)
gdsfile <- "1KG_phase3_subset_chr1.gds"
if (!file.exists(gdsfile)) download.file(file.path(data.path, gdsfile), gdsfile)
gds <- seqOpen(gdsfile)
seqData <- SeqVarData(gds, sampleData=annot)
assoc <- assocTestMM(seqData, nullmod)
head(assoc)
```

We make a QQ plot to examine the results.

```{r assoc_single_qq}
library(ggplot2)
qqPlot <- function(pval) {
    pval <- pval[!is.na(pval)]
    n <- length(pval)
    x <- 1:n
    dat <- data.frame(obs=sort(pval),
                      exp=x/n,
                      upper=qbeta(0.025, x, rev(x)),
                      lower=qbeta(0.975, x, rev(x)))
    
    ggplot(dat, aes(-log10(exp), -log10(obs))) +
        geom_line(aes(-log10(exp), -log10(upper)), color="gray") +
        geom_line(aes(-log10(exp), -log10(lower)), color="gray") +
        geom_point() +
        geom_abline(intercept=0, slope=1, color="red") +
        xlab(expression(paste(-log[10], "(expected P)"))) +
        ylab(expression(paste(-log[10], "(observed P)"))) +
        theme_bw()
}    

qqPlot(assoc$Wald.pval)
```

## Sliding window tests

For rare variants, we can do burden tests or SKAT on sliding windows using the GENESIS function `assocTestSeqWindow`. We restrict the test to variants with alternate allele frequency < 0.1. (For real data, this threshold would be lower.) We use a flat weighting scheme.

```{r assoc_window_burden}
assoc <- assocTestSeqWindow(seqData, nullmod, test="Burden", AF.range=c(0,0.1),
                            weight.beta=c(1,1), window.size=5, window.shift=2)
names(assoc)
head(assoc$results)
head(assoc$variantInfo)

qqPlot(assoc$results$Score.pval)
```

For SKAT, we use the Wu weights.

```{r assoc_window_skat}
assoc <- assocTestSeqWindow(seqData, nullmod, test="SKAT", AF.range=c(0,0.1),
                            weight.beta=c(1,25), window.size=5, window.shift=2)
head(assoc$results)
head(assoc$variantInfo)

qqPlot(assoc$results$pval_0)
```

## Exercises

1. Logistic regression: `fitNullMM` can use a binary phenotype as the outcome variable by specifying the argument `family=binomial`. Use the `status` column in the sample annotation to fit a null model for simulated case/control status. Refer to the documentation for `fitNulMM` to see what other parameters need to be changed for a binary outcome. Then run a single-variant test and a sliding window test using this model.

```{r exercise_logistic, include=FALSE, eval=FALSE}
nullmod <- fitNullMM(annot, outcome="status", covars=c("sex", "Population"), 
                     covMatList=grm$grm, family=binomial)
assoc <- assocTestMM(seqData, nullmod, test="Score")
assoc <- assocTestSeqWindow(seqData, nullmod, test="Burden", AF.range=c(0,0.1),
                            weight.beta=c(1,1), window.size=5, window.shift=2)
```

2. Inverse normal transform: use the TopmedPipeline function `addInvNorm` to perform an inverse normal transform on the `status` variable. (You will need to load the TopmedPipeline library.) Inspect the code for this function by typing `addInvNorm` at the R prompt, so you understand what it is doing.


## Aggregate tests

### Variant annotation
Rare variants are generally aggregated into some meaningful units for association testing to decrease multiple testing burden and increase statistical power. Various genomic and epigenomic annotations can be used to define aggregation units and filter them. A large number of annotations are available through the Whole Genome Sequence Annotator (WGSA) to the TOPMed users. 

### Defining aggregate units
We will be using a gene-based aggregation unit, where each unit is a GENCODE gene and 20 kb flanking region upstream and downstream of it. For real data, one will likely filter variants within each unit based on various annotations (examples include loss of function, conservation, deleteriousness scores, etc.).

The aggregation units are defined in an R dataframe. Each row of the dataframe specifies a variant (chromosome, position, ref, alt) and the group identifier (group_id) assigned to it. Mutiple rows with different group identifiers can be specified to assign a variant to different groups (for example a variant can be assigned to mutiple genes).

```{r agg_unit}
aggfile <- "variants_by_gene.RData"
if (!file.exists(aggfile)) download.file(file.path(workshop.path, aggfile), aggfile)
aggunit <- TopmedPipeline::getobj(aggfile)
names(aggunit)
head(aggunit)

# an example of variant that is present in mutiple groups
library(dplyr)
mult <- aggunit %>%
    group_by(chromosome, position) %>%
    summarise(n=n()) %>%
    filter(n > 1)
inner_join(aggunit, mult[2,1:2])
```

### Association testing with aggregate units

We can run a burden test or SKAT on each of these units using the GENESIS function `assocTestSeq`. This function expects a list, where each element of the list is a dataframe representing a single aggregation unit and containing the unique variant.id assigned to each variant in a GDS file. We use the TopmedPipeline function `aggregateListByAllele` to quickly convert our single dataframe to the required format. This function can account for multiallelic variants (the same chromosome, position, and ref, but different alt alleles). The first argument is the GDS object returned by `seqOpen` (see above).

```{r aggVarList}
library(TopmedPipeline)
aggVarList <- aggregateListByAllele(gds, aggunit)
length(aggVarList)
head(names(aggVarList))
aggVarList[[1]]
```

As in the previous section, we must fit the null model before running the association test.

```{r assoc_aggregate}
assoc <- assocTestSeq(seqData, nullmod, test="Burden", aggVarList=aggVarList, 
                      AF.range=c(0,0.1), weight.beta=c(1,1))
names(assoc)
head(assoc$results)
head(names(assoc$variantInfo))
head(assoc$variantInfo[[1]])

qqPlot(assoc$results$Score.pval)
```

```{r assoc_close}
seqClose(gds)
```


### Exercise

Since we are working with a subset of the data, many of the genes listed in `group_id` have a very small number of variants. Create a new set of units based on position rather than gene name, using the TopmedPipeline function `aggregateListByPosition`. Then run SKAT using those units.

```{r exercise_aggregate, include=FALSE, eval=FALSE}
agg2 <- aggunit %>%
    mutate(chromosome=factor(chromosome, levels=c(1:22, "X"))) %>%
    select(chromosome, position) %>%
    distinct() %>%
    group_by(chromosome) %>%
    summarise(min=min(position), max=max(position))
aggByPos <- bind_rows((lapply(1:nrow(agg2), function(i) {
    data.frame(chromosome=agg2$chromosome[i],
               start=seq(agg2$min[i], agg2$max[i]-1e6, length.out=10),
               end=seq(agg2$min[i]+1e6, agg2$max[i], length.out=10))
})) %>%
    mutate(group_id=1:n())

aggVarList <- aggregateListByPosition(gds, aggByPos)
assoc <- assocTestSeq(seqData, nullmod, test="SKAT", aggVarList=aggVarList,
                      AF.range=c(0,0.1), weight.beta=c(1,25))
head(assoc$results)
```
