# GDS format

GDS is Genomic Data Structure, a storage format that can efficiently store genomic data and provide fast random access to subsets of the data. For more information on GDS for sequence data, read the [SeqArray package vignette](https://github.com/zhengxwen/SeqArray/blob/master/vignettes/SeqArrayTutorial.Rmd).

## Exploring a GDS file

To use the R packages developed at the DCC for sequence data, we first need to convert a VCF file to GDS. (If the file is BCF, use [https://samtools.github.io/bcftools/bcftools.html](bcftools) to convert to VCF.)

```{r vcf2gds}
library(SeqArray)
data.path <- "https://github.com/UW-GAC/analysis_pipeline/raw/master/testdata"
vcffile <- "data/1KG_phase3_subset_chr1.vcf.gz"
if (!dir.exists("data")) dir.create("data")
if (!file.exists(vcffile)) download.file(file.path(data.path, basename(vcffile)), vcffile)
gdsfile <- "data/1KG_phase3_subset_chr1.gds"
seqVCF2GDS(vcffile, gdsfile, fmt.import="GT", storage.option="LZMA_RA", verbose=FALSE)
```

We can interact with the GDS file using the SeqArray package.

```{r seqarray}
gds <- seqOpen(gdsfile)
gds

# the unique sample identifier comes from the VCF header
sample.id <- seqGetData(gds, "sample.id")
length(sample.id)
head(sample.id)

# a unique integer ID is assigned to each variant
variant.id <- seqGetData(gds, "variant.id")
length(variant.id)
head(variant.id)

# reference allele frequency of each variant
afreq <- seqAlleleFreq(gds)
hist(afreq, breaks=50)
```

We can define a filter on the `gds` object. After using the `seqSetFilter` command, all subsequent reads from the `gds` object are restricted to the selected subset of data, until a new filter is defined or `seqResetFilter` is called.

```{r filter}
seqSetFilter(gds, variant.id=1:10, sample.id=sample.id[1:5])
```

Genotype data is stored in a 3-dimensional array, where the first dimension is always 2 for diploid genotypes. The second and third dimensions are samples and variants, respectively. The values of the array denote alleles: `0` is the reference allele and `1` is the alternate allele. For multiallelic variants, other alternate alleles are represented as integers `> 1`.

```{r genotypes}
geno <- seqGetData(gds, "genotype")
dim(geno)
geno[,,1:2]
```

The [SeqVarTools package](http://bioconductor.org/packages/SeqVarTools) has some additional functions for interacting with SeqArray-format GDS files.

```{r seqvartools}
library(SeqVarTools)

# return genotypes in matrix format
getGenotype(gds)
getGenotypeAlleles(gds)
refDosage(gds)
altDosage(gds)

# look at reference and alternate alleles
refChar(gds)
altChar(gds)

# reset the filter to all variants and samples
seqResetFilter(gds)

# how many alleles for each variant?
n <- seqNumAllele(gds)
table(n)

# some variants have more than one alternate allele
multi.allelic <- which(n > 2)
altChar(gds)[multi.allelic]

# extract a particular alternate allele
# first alternate
altChar(gds, n=1)[multi.allelic]
# second alternate
altChar(gds, n=2)[multi.allelic]

# how many variants are SNVs vs INDELs?
table(isSNV(gds, biallelic=TRUE))
table(isSNV(gds, biallelic=FALSE))
# 11 SNVs are multi-allelic
```

## Exercises 

1. Set a filter selecting only multi-allelic variants. Inspect their genotypes using the different methods you learned above. Use the `alleleDosage` method to find dosage for the second (and third, etc.) alternate allele.

```{r exercise_gds, include=FALSE}
seqSetFilter(gds, variant.sel=multi.allelic)

geno <- seqGetData(gds, "genotype")
dim(geno)
geno[,1:5,]

geno <- getGenotype(gds)
dim(geno)
head(geno)

geno <- getGenotypeAlleles(gds)
head(geno)

dos <- refDosage(gds)
head(dos)

dos <- altDosage(gds)
head(dos)

dos <- alleleDosage(gds, n=2)
head(dos)

dos <- alleleDosage(gds, n=3)
head(dos)
```

2. Use the `hwe` function in SeqVarTools to run a Hardy-Weinberg Equilibrium test on each variant. Identify a variant with low p-value and inspect its genotypes. (Note that the HWE test is only valid for biallelic variants, and will return `NA` for multiallelic variants.)

```{r exercise_hwe, include=FALSE}
seqResetFilter(gds)
hwe.res <- hwe(gds)
lowp <- !is.na(hwe.res$p) & hwe.res$p < 1e-4
head(hwe.res[lowp,])

seqSetFilter(gds, variant.id=75)
table(getGenotype(gds))
table(refDosage(gds))
```

```{r intro_close}
seqClose(gds)
```
