# Phenotype Harmonization

Because TOPMed comprises multiple, distinct studies, it is necessary to perform phenotype harmonization before running a cross-study analysis.
Harmonization is generally performed by "harmonization unit", which is defined as a group of subjects whose phenotypes can be similarly processed.
In many cases, one study corresponds to one harmonization unit, but more complicated studies may require multiple harmonization units.
For example, the Framingham Heart study has multiple subcohorts (Original, Offspring, etc.), with phenotypes measured differently for subjects in different cohorts.
Since the phenotypes have been measured differently, the different subcohorts have to be harmonized separately.

## Exercise overview: run diagnostics on a harmonized phenotype

In this exercise, we assume that you have
created a phenotype harmonization plan for height,
sent it to members from three studies to perform the harmonization,
and
received a harmonized phenotype file from each study.
We will generate some diagnostic information about the harmonized phenotype.

The exercise uses 1000 Genomes data, with simulated phenotypes for study, age, and height.

## Import phentoype files into R

The first step is to read the files into R for processing.
Before we begin, you need to download the data from github so you have access to it.

```{r}
repo_path <- "https://github.com/UW-GAC/topmed_workshop_2017/raw/master"
pheno_files <- c("pheno_data_study_1.txt", "pheno_data_study_2.txt", "pheno_data_study_3.txt")
for (pheno_file in pheno_files) {
  if (!file.exists(pheno_file)) download.file(file.path(repo_path, pheno_file), pheno_file)
}
```

Next, read the study phenotype files into R.
In this case, each file is tab-delimited and has the same phenotype variable names.

```{r}
study_1 <- read.table("pheno_data_study_1.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_1)

study_2 <- read.table("pheno_data_study_2.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_2)

study_3 <- read.table("pheno_data_study_3.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_3)
```

We will be looking at differences by harmonization unit (in this case, study), so add the study identifier to the data frame.
```{r}
study_1$study <- "study_1"
study_2$study <- "study_2"
study_3$study <- "study_3"
```

Combine the three different study data frames into one large data frame for joint analysis.
Before doing this, we should check that the column headers are the same.
```{r, message = FALSE}
all.equal(names(study_1), names(study_2))
all.equal(names(study_1), names(study_3))

library(dplyr)
phen <- dplyr::bind_rows(study_1, study_2, study_3)
```


## Run some quick diagnostics on the files

We can look at the distribution of phenotype data with text-based reports or with plots.

First, inspect distributions with `table` for categorical traits and with `summary` for quantitative traits.
The commads are shown here for study_1, but you should run them for study_2 and study_3 as well to see if you can see any differences.
```{r}
table(study_1$sex)

summary(study_1$age)

summary(study_1$height)
```

It is also helpful to use plots to inspect the distributions of phenotype data.
Here, we will look at boxplots of height by study.
```{r, message = FALSE}
library(ggplot2)
ggplot(phen, aes(x = study, y = height)) + geom_boxplot()
```

You may also want to see the difference in height when you include both study and sex:
```{r}
ggplot(phen, aes(x = study, fill = sex, y = height)) + geom_boxplot()
```

These diagnostics are helpful to get a feel for the data.
They can help you see if one study is vastly different from the others or detect outlier values.
Some of the differences could be accounted for by covariates, so we will assess the statistical significance of study differences in the next section.


## Fit a mixed model for statistical comparisons

The quick diagnostics in the previous section let you see if the data from one study are completely different from the others, but they don't provide much statistical insight.
For that, we need to fit a statistical model to the data.
Because some of the studies in TOPMed have related individuals, we need to fit a mixed model to account for the correlation in the data.
In this case, because the phenotype is quantitative, we will use a linear mixed model.
More information about mixed models will be given during presentations tomorrow.

We use the `GENESIS` R package for fitting the mixed model.
This package can accept a correlation matrix as a random effect in the mixed model, insetad of requiring a categorical or indicator variable.
It therefore can account for the observed genetic relatedness between subjects.
It is also the same package that we use for the association analyses, so this exercise provides a brief introduction to the package and some of the associated data structures.

### Create an Annotated Data Frame

The first step in fitting the mixed model is to create an Annotated Data Frame.
This data structure is provided by the Bioconductor `Biobase` package, and it contains both the data and metadata (primarily the variable description).

The `GENESIS` code to fit the mixed model requires a `scanID` column.
Typically the `scanID` column identifies a unique genotyping instance and is linked to a sample id, not a subject id.
In this case, we are only working with subject-level data, so we can use the subject identifier as the scan identifier for model-fitting purposes.
```{r}
phen$scanID <- phen$subject_id
```

Next, create the Annotated Data Frame.
You should include a description of each variable in the metadata.
```{r, message = FALSE}
library(Biobase)

metadata <- data.frame(labelDescription = c(
  "subject identifier",
  "subject's sex",
  "age at measurement of height",
  "subject's height in cm",
  "study identifier",
  "scan identifier"
))

annot <- AnnotatedDataFrame(phen, metadata)

# access the data with the pData() function
head(pData(annot))

# access the metadata with the varMetadata() function
varMetadata(annot)
```


### Fit the mixed model

Becase it is an input to the mixed model, we next need to download the genetic relatedness matrix calculated for these subjects.
```{r}
data_path <- "https://github.com/smgogarten/analysis_pipeline/raw/devel/testdata"
grmfile <- "grm.RData"
if (!file.exists(grmfile)) download.file(file.path(data_path, grmfile), grmfile)
grm <- TopmedPipeline::getobj(grmfile)
rownames(grm$grm) <- colnames(grm$grm) <- grm$sample.id
```

Now, set up the model.
We will use `height` as the outcome variable and adjust for `sex`, `age`, and harmonzation unit (here, `study`) as fixed effects.
We will also include the genetic relatedness matrix as a random effect.
The code also requires the AnnotatedDataFrame scan identifiers to be in the same order as the samples in the genetic relatedness matrix.
````{r, message = FALSE}
library(GENESIS)

outcome <- "height"
covars <- c("sex", "age", "study")

covariance_matrices <- grm$grm

# put the phenotype data in the same order as the GRM
annot <- annot[match(grm$sample.id, annot$scanID), ]
```


Next, we will fit a model that allows for group-specific residual variance by harmonization unit (here, `study`) by using the `group.var` argument of `fitNulLMM`.
```{r}
mod <- GENESIS::fitNullMM(annot, outcome = outcome, covars = covars, group.var = "study",
                     covMatList = covariance_matrices, verbose = FALSE)
```

The output of `fitNullMM` is a list with a number of components.
```{r}
names(mod)
```

The elements that we will work with in this exercise are:

* `converged`: an indicator of whether the model successfully converged
* `model.matrix`: The matrix of subject-covariate values used to fit the model
* `fixef`: The fitted fixed effects
* `betaCov`: The covariance of the fitted fixed effects
* `resid.marginal`: The (marginal) residuals from the model, which have been adjusted for the fixed effects but not for the covariance structure
* `varComp`: The fitted variance component for each input covariance matrix
* `logLik`: The log-likelihood of the model fit


### Model interpretation

First, add the residuals to the phenotype data frame for plotting later in this section.
We need to make sure that we are matching each residual value to the correct subject.
To do this, we use the row names of the `model.matrix` element of the output, which are in the same order as the `residual` matrix.
We then match the row names (and therefore the residuals) to the order of the subjects in the phenotype file using the base R function `match`.
```{r}
j <- match(phen$subject_id, rownames(mod$model.matrix))
phen$residuals <- mod$resid.marginal[j]
```


Next, we want to check if the different studies have the same mean height after adjustment for other covariates (here, age and sex).
This information is available in the `fixef` (fixed effects) element of the model output.

```{r}
mod$fixef
```

In particular, the rows associated with the study effects have significant estimates associated with them.

We can also run a Wald test, where the null hypothesis is that all of the studies have the same mean height.
First, identify the fixed effects and covariance for the study indicator variables.
```{r}
# these rows are associated with the study effects
idx <- 4:5

# fixed effect estimates for the studies
study_effect <- mod$fixef[idx, ]
study_effect

# covariance of the study effects
study_cov <- mod$betaCov[idx, idx]
study_cov
```

Next, calculate the Wald test statistic and associated p-value.
```{r}
test_stat <- t(study_effect$Est) %*% solve(study_cov) %*% study_effect$Est
pval <- pchisq(test_stat, df = length(study_effect), lower.tail = FALSE)
pval
```

The small p-value indicates that the null hypothesis is rejected; the height measurements from the studies do not have the same mean height even after adjusting for the other covariates.


Finally, we want to check if the height distributes from the different studies have the same variance.

Start by looking at the variance components `varComp` element of the model.
```{r}
mod$varComp
```

`V_A` represents the variance in height due to genetic relatedness.
The other variance components (`V_study_1`, `V_study_2`, and `V_study_3`) represent the residual variance in each study.
The fitted values of the variance components are different for the different studies, indicating that the distribution of height in the three studies has different variance even after accounting for the other covariates.

We can also show this same information by plotting the residuals by study.
```{r}
ggplot(phen, aes(x = study, y = residuals)) +
  geom_boxplot()
```

Both methods of looking at the variance components indicate that study 1 has a much lower residual variance than the others.
It is more complicated to perform a test to see if the difference is statistically significant.
The distribution of the null hypothesis is not a simple chi-square distribution due to the fact that variance components cannot be negative.
Therefore, we will only discuss the qualitative differences in variance between the studies at this point.
The observed difference in variance may be due to a biological reason, or it might be due to differences in measurement that might disqualify study 1 from inclusion in the combined analysis.


## Final considerations

We have determined that the different studies have both different mean and different variance by study for height.
Before performing genotype-phenotype association tests with these data, you would need to think carefully about whether the phenotype is homogeneous enough to be analyzed together.
In some cases, there may be a valid reason for different means or variances, for example:

* different heights in different study populations, such as a study composed primarily of Asian participants vs. a study with primarily European participants
* possible secular trends in height, such as comparing the Framingham Original cohort from ~1950 to a cohort from the present day

In other cases, there may be good reasons to exclude one or more studies, for example:

* a systematic measurement error in one study
* miscalculation or misinterpretation of the harmonization algorithm
* study populations that are too different to be compared, such as trying to include a study composed primarily of children with one composed of adults in a height analysis

Unfortunately there is no single set of guidelines you can use to make this decision, so it requires careful consideration before proceeding with analyses.
It is necessary to involve domain experts (e.g., the Working Group members) and study experts to determine whether the phenotype is homogeneous enough to be analyzed together.
