# Phenotype Harmonization

Because TOPMed comprises multiple, distinct studies, it is necessary to perform phenotype harmonization before running a cross-study analysis.
Harmonization is generally performed by "harmonization unit", which is defined as a group of subjects whose phenotypes can be similarly processed.
In many cases, one study corresponds to one harmonization unit, but more complicated studies may require multiple harmonization units.
For example, the Framingham Heart study has multiple subcohorts (Original, Offspring, etc.), with phenotypes measured differently for subjects in different cohorts.
Since the phenotypes have been measured differently, the different subcohorts have to be harmonized separately.

## Exercise overview: run diagnostics on a harmonized phenotype

In this exercise, we assume that you have
created a phenotype harmonization plan for height,
sent it to members from three studies to perform the harmonization,
and
received a harmonized phenotype file from each study.
We will generate some diagnostic information about the harmonized phenotype.

The exercise uses 1000 Genomes data, with simulated phenotypes for sex, age, and height.

## Import phentoype files into R

The first step is to read the files into R for processing.
Before we begin, you need to download the data from github so you have access to it.

```{r}
repo_path <- "https://github.com/UW-GAC/topmed_workshop_2017/"
pheno_files <- c("pheno_data_study_1.txt", "pheno_data_study_2.txt", "pheno_data_study_3.txt")
for (pheno_file in pheno_files) {
  if (!file.exists(pheno_file)) {
    download.file(file.path(pheno_data_path, pheno_file), pheno_file)
  }
}
```

Next, read the study phenotype files into R.
In this case, each file is tab-delimited and has the same phenotype variable names.

```{r}
study_1 <- read.table("pheno_data_study_1.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_1)

study_2 <- read.table("pheno_data_study_2.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_2)

study_3 <- read.table("pheno_data_study_3.txt", header = TRUE, sep = "\t", as.is = TRUE)
head(study_3)
```

We will be looking at differences by harmonization unit (in this case, study), so add the study identifier to the data frame.
```{r}
study_1$study <- "study_1"
study_2$study <- "study_2"
study_3$study <- "study_3"
```

Combine the three different study data frames into one large data frame for joint analysis.
Before doing this, we should check that the column headers are the same.
```{r, message = FALSE}
all.equal(names(study_1), names(study_2))
all.equal(names(study_1), names(study_3))

library(dplyr)
phen <- dplyr::bind_rows(study_1, study_2, study_3)
```


## Run some quick diagnostics on the files

We can look at the distribution of phenotype data with text-based reports or with plots.

First, inspect distributions with `table` for categorical traits and with `summary` for quantitative traits.
The commads are shown here for study_1, but you should run them for study_2 and study_3 as well to see if you can see any differences.
```{r}
table(study_1$sex)

summary(study_1$age)

summary(study_1$height)
```

It is also helpful to use plots to inspect the distributions of phenotype data.
Here, we will look at boxplots of height by study.
```{r, message = FALSE}
library(ggplot2)
ggplot(phen, aes(x = study, y = height)) + geom_boxplot()
```

You may also want to see the difference in height when you include both study and sex:
```{r}
ggplot(phen, aes(x = study, fill = sex, y = height)) + geom_boxplot()
```

These diagnostics are helpful to get a feel for the data.
They can help you see if one study is vastly different from the others or detect outlier values.
Some of the differences could be accounted for by covariates, so we will look at how to assess the statistical significance of study differences in the next section.


## Fit a mixed model for statistical comparisons

The quick diagnostics in the previous section let you see if the data from one study are completely different from the others, but they don't provide much statistical insight.
For that, we need to fit a statistical model to the data.
Because some of the studies in TOPMed have related individuals, we need to fit a mixed model to account for the correlation in the data.
In this case, because the phenotype is quantitative, we will use a linear mixed model.
More information about mixed models will be given during presentations tomorrow.

We use the `GENESIS` R package for fitting the linear model.
This package can accept a correlation matrix as a random effect in the linear model, insetad of requiring indicator variables.
It therefore can account for the observed genetic relatedness between subjects.
It is also the same package that we use for the association analyses, so this exercise provides a brief introduction to the package and some of the associated data structures.

### Create an annotated data frame

The first step in fitting the linear model is to create an Annotated Data Frame.
This data structure is provided by the Bioconductor `Biobase` package, and it contains both the data and metadata (primarily the variable description).

The `GENESIS` code to fit the linear model requires a `scanID` column.
Typically the `scanID` column identifies a unique genotyping instance and is linked to a sample id, not a subject id.
In this case, we are only working with subject-level data, so we can use the subject identifier as the scan identifier for model-fitting purposes.
```{r}
phen$scanID <- phen$subject_id
```

Next, create the annotated data frame.
You should include a description of each variable in the metadata.
```{r, message = FALSE}
library(Biobase)

metadata <- data.frame(labelDescription = c(
  "scan identifier",
  "subject identifier",
  "subject's sex",
  "study identifier",
  "age at measurement of height",
  "subject's height in cm"
))

annot <- AnnotatedDataFrame(phen, metadata)

# access the data with the pData() function
head(pData(annot))

# access the metadata with the varMetadata() function
varMetadata(annot)
```


### Fit the linear mixed model

Becase it is an input to the linear mixed model, we next need to download the genetic relatedness matrix calculated for these subjects.
```{r}
grmfile <- "grm.RData"
if (!file.exists(grmfile)) download.file(file.path(data.path, grmfile), grmfile)
grm <- TopmedPipeline::getobj(grmfile)
rownames(grm$grm) <- colnames(grm$grm) <- grm$sample.id
```

Now, set up the model.
We will use `height` as the outcome variable and adjust for `sex`, `age`, and harmonzation unit (here, `study`) as fixed effects.
We will also include the genetic relatedness matrix as a random effect.
The code also requires the AnnotatedDataFrame scan identifiers to be in the same order as the samples in the genetic relatedness matrix.
````{r, message = FALSE}
library(GENESIS)

outcome <- "height"
covars <- c("sex", "age", "study")

covariance_matrices <- grm$grm

# put the phenotype data and the GRM in the same order
annot <- annot[match(grm$sample.id, annot$scanID), ]
```


Next, we will fit two models:

1. Homoskedastic - do not allow different variances by harmonization unit
2. Heteroskedastic - allow for difference variances by harmonization unit

Both models adjust for the same set of covariates and account for the same random effects.
To fit the heteroskedastic model, we use the `group.var` argument to specify a column in the Annotated Data Frame to use to fit group-specific variances.
In this case, we will allow the different studies to have different fitted variances.
```{r}
nullmod_hom <- fitNullMM(annot, outcome = outcome, covars = covars,
                     covMatList = covariance_matrices, verbose = FALSE)
nullmod_het <- fitNullMM(annot, outcome = outcome, covars = covars, group.var = "study",
                     covMatList = covariance_matrices, verbose = FALSE)
```

The output of `fitNullMM` is a list with a number of components.
```{r}
names(nullmod_hom)
```

The elements that we will work with in this exercise are:

* `varComp`: The fitted variance component for each input covariance matrix
* `fixef`: The fitted fixed effects
* `betaCov`: The covariance of the fitted fixed effects
* `resid.marginal`: The (marginal) residuals from the model, which have been adjusted for the fixed effects but not for the covariance structure
* `logLik`: The log-likelihood of the model fit
* `hetResid`: an indicator of whether the model was heteroskedastic
* `converged`: an indicator of whether the model successfully converged
* `model.matrix`: The matrix of subject-covariate values used to fit the model

### Model interpretation

The fixed effects from the models should be roughly the same, with small differences only due to the iterative fitting algorithm:
```{r}
nullmod_hom$fixef

nullmod_het$fixef
```

However, the variance components are different.
In both models, `V_A` represents the variance due to genetic relatedness.
In the homoskedastic model, `V_E` represents the remainder of variance; there is only one value.
On the other hand, the heteroskedastic model has three residual variance components - one for each study.
In this case, the variance components are different for different studies, indicating that the distribution of height in the three studies has different variance even after accounting for the other parameters.
```{r}
nullmod_hom$varComp

nullmod_het$varComp
```

We can run a likelihood ratio test to check this hypothesis statistically.
Because the models are identical other than the number of variance components fit, the heterogeneous model has 2 more degrees of freedom.
We use the likelihood ratio difference and the additional degrees of freedom to calculate a p-value.
```{r}
D <- 2 * (nullmod_het$logLik - nullmod_hom$logLik)
pchisq(D, df = 2, lower.tail = FALSE)
```
Here, a small p-value means that study-specific variances are significantly different from each other.
In this case, the heterogeneous model is preferred, so we will work with it for the future checks.
If you were to run an association analysis with this phenotype, it would be important to use a heteroskedastic model to account for the different variances by study.

Now, add the residuals to the phenotype data frame for plotting.
We first need to make sure that we are matching the each residual value to the correct subject.
To do this, we use the row names of the `model.matrix` element of the output, which are in the same order as the `residual` matrix.
We then match them to the order of the subjects in the phenotype file using the base R command `match`.
```{r}
j <- match(phen$subject_id, rownames(nullmod_het$model.matrix))
phen$residuals <- nullmod_het$resid.marginal[j]
```


We start by plotting the distribution of the residuals by study.
```{r}
ggplot(phen, aes(x = study, y = residuals)) +
  geom_violin() +
  geom_boxplot(width = 0.1)
```

We had earlier seen that the likelihood ratio test indicated that the variance was different for the studies.
Here we can see that `study_1` has a smaller variance than the others.

We also want to check if the different studies have different means.
This information is available in the fixed effects part of the null model, which we briefly looked at earlier.
```{r}
nullmod_het$fixef
```
In particular, the rows associated with study effects have significant estimates associated with them.

We can also run a Wald test, where the null hypothesis is that all of the studies have the same mean height.
First, identify the fixed effects and covariance for the study indicator variables.
```{r}
# these rows are associated with the study effects
idx <- 4:5

# fixed effect estimates for the studies
study_effect <- nullmod_het$fixef[idx, ]
study_effect

# covariance of the study effects
study_cov <- nullmod_het$betaCov[idx, idx]
study_cov
```

Next, calculate the Wald test statistic and associated p-value.
```{r}
test_stat <- t(study_effect$Est) %*% solve(study_cov) %*% study_effect$Est
pval <- pchisq(test_stat, df = length(study_effect), lower.tail = FALSE)
pval
```

The small p-value indicates that the null hypothesis is rejected; the studies do not have the same mean height.

## Final considerations

We have determined that the different studies have both different variance and different mean for height.
Before performing genotype-phenotype association tests with these data, you would need to think carefully about whether the phenotype is homogeneous enough to be analyzed together.
In some cases, there may be a valid reason for different means or variances, for example:

* different heights in different study populations, such as a study composed primarily of Asian participants vs. a study with primarily European participants
* possible secular trends in height, such as comparing the Framingham Original cohort from ~1950 to a cohort from the present day

In other cases, there may be good reasons to exclude one or more studies, for example:

* a systematic measurement error in one study
* miscalculation or misinterpretation of the harmonization algorithm
* study populations that are too different to be compared, such as trying to include a study composed primarily of children with one composed of adults in a height analysis

Unfortunately there is no single set of guidelines you can use to make this decision, so it requires careful consideration before proceeding with analyses.
It is necessary to involve domain experts (e.g., the Working Group members) and study experts to determine whether the phenotype is homogeneous enough to be analyzed together.
